---
title: "Week 4 Lab: Who Does Indego Reach?"
subtitle: "MUSA 5080 · In-class team challenge (ungraded)"
format:
  html:
    toc: true
    code-fold: false
execute:
  freeze: false
  warning: false
  message: false
---

## How this works

This is a **Lab**: it's ungraded and done in class. You work in teams of 3–4.

**Type the code. Don't copy and paste it.** Blanks look like `___`. Fill them in, then run the chunk.

The first hour (Parts 1–5) is guided and builds your toolkit. The last part is the **challenge**: your team answers one policy question and defends one decision. At the end, the presenter from each team is chosen at random, so everyone needs to understand every step.

> **The question:** Indego says it serves low-income Philadelphians.
> Which low-income tracts are farthest from a station, and how sure are we that they're low-income?

---

## Part 0: Setup

```{r}
#| label: setup
library(tidyverse)
library(tidycensus)
library(sf)

# Your key lives in .Renviron. Never paste it here.
# tidycensus reads CENSUS_API_KEY automatically.
Sys.getenv("CENSUS_API_KEY") != ""   # should print TRUE
```

---

## Part 1: The same pull as Week 3, plus one argument

This is the same `get_acs()` call you already know. The only new part is `geometry = TRUE`.

```{r}
#| label: pull-tracts
philly_tracts <- get_acs(
  geography = "tract",
  state = "PA",
  county = "Philadelphia",
  variables = c(med_inc = "B19013_001",
                total_pop = "B01003_001"),
  year = 2023,
  output = "wide",
  geometry = TRUE
)

class(philly_tracts)
glimpse(philly_tracts)
```

**Stop and look:** What does `class()` show? Which column is new?

Now repeat the Week 3 steps. **A data frame goes in, a data frame comes out.** The geometry column comes along automatically.

```{r}
#| label: reliability
philly_tracts <- philly_tracts %>%
  filter(total_popE > 0) %>%
  mutate(
    med_inc_se = med_incM / 1.645,
    med_inc_cv = (med_inc_se / med_incE) * 100,
    reliability = case_when(
      med_inc_cv < 12  ~ "High",
      med_inc_cv <= 40 ~ "Moderate",
      TRUE             ~ "Low"
    )
  )

# Did geometry survive? It should.
names(philly_tracts)
```

Your first map. It's just ggplot with a new geom:

```{r}
#| label: first-map
ggplot(philly_tracts) +
  geom_sf(aes(fill = med_incE), color = NA) +
  scale_fill_viridis_c(labels = scales::dollar) +
  theme_void() +
  labs(title = "Median household income, Philadelphia tracts",
       fill = NULL)
```

**Quick check:** How many tracts have `NA` income even though they have people living in them? Write the code to count them.

```{r}
#| label: na-check
philly_tracts %>%
  filter(is.na(med_incE)) %>%
  nrow()
```

---

## Part 1b: Join by what

How much has income changed since 2015? Pull the older estimates. No geometry is needed, because this is just a table.

```{r}
#| label: pull-2015
income_15 <- get_acs(
  geography = "tract",
  state = "PA",
  county = "Philadelphia",
  variables = c(med_inc_15 = "B19013_001"),
  year = 2015,
  output = "wide",
  geometry = FALSE
) %>%
  select(GEOID, med_inc_15E, med_inc_15M)

nrow(income_15)
nrow(philly_tracts)
```

**Stop and look:** The row counts are different. Before you join, predict why.

Now join the 2015 table onto your tracts. The key is the column both tables share.

```{r}
#| label: left-join
philly_tracts <- philly_tracts %>%
  left_join(income_15, by = "GEOID")

nrow(philly_tracts)   # did the row count change?
```

**The rule: after every join, count rows and count NAs.**

```{r}
#| label: join-check
philly_tracts %>%
  st_drop_geometry() %>%
  summarize(
    n_tracts = n(),
    no_2015_match = sum(is.na(med_inc_15E.x))
  )
```

Now look at which tracts failed to match:

```{r}
#| label: join-failures
# 2023 tracts with no 2015 partner
philly_tracts %>%
  st_drop_geometry() %>%
  filter(is.na(med_inc_15E.x)) %>%
  select(GEOID, NAME) %>%
  head(10)

# 2015 tracts with no 2023 partner: anti_join keeps rows with NO match
income_15 %>%
  anti_join(st_drop_geometry(philly_tracts), by = "GEOID") %>%
  head(10)
```

**Discuss as a team:** Did these places move? What changed between 2015 and 2023? Why can't `left_join()` fix this? Keep your answer in mind; Part 3 comes back to it.

::: {.callout-note}
Some NAs aren't boundary changes. A tract can match but have `NA` income because the Census suppressed the estimate. How would you tell the two kinds of NA apart?
:::

---

## Part 2: Bring in the stations

Indego publishes live station locations as GeoJSON. `st_read()` reads it directly from the URL.

```{r}
#| label: stations
stations <- st_read("https://www.rideindego.com/stations/json/", quiet = TRUE)

nrow(stations)
glimpse(stations)
```

Now check the CRS of both layers.

```{r}
#| label: crs-check
st_crs(philly_tracts)$epsg
st_crs(___)$epsg
```

**Predict before you run:** The two numbers are different. What will happen if you try to join them?

```{r}
#| label: crs-fail
#| error: true
st_join(stations, philly_tracts)
```

That error is sf protecting you. Fix it by **translating** both layers into Pennsylvania South State Plane, EPSG 2272.

::: {.callout-warning}
## Label or translation?
`st_set_crs()` writes a label when the CRS is **missing**. `st_transform()` converts coordinates into a different CRS. Both layers here already have a CRS. Which function do you need?
:::

```{r}
#| label: transform
philly_tracts <- philly_tracts %>% ___(crs = 2272)
stations      <- stations %>% ___(crs = 2272)

st_crs(philly_tracts)$units   # what units are we in now?
```

**Write down the units.** Every number you type from now on is in these units.

---

## Part 3: Join by where

In Part 1b, the join failed because it matched on a label (GEOID), and the label changed. `st_join()` matches on **location** instead: a row is matched to whatever it sits on top of. Stations don't have a GEOID column, and they don't need one.

Every station gets its tract's attributes:

```{r}
#| label: spatial-join
stations_with_tract <- st_join(stations, philly_tracts)

stations_with_tract %>%
  st_drop_geometry() %>%
  select(name, GEOID, med_incE, reliability) %>%
  head()
```

Now go the other way and count stations **per tract**. The first route is the Week 2 collapse pattern:

```{r}
#| label: count-route1
station_counts <- stations_with_tract %>%
  st_drop_geometry() %>%
  ___(GEOID)
```

The second route counts directly:

```{r}
#| label: count-route2
philly_tracts <- philly_tracts %>%
  mutate(n_stations = lengths(st_intersects(., stations)))
```

### Check your work

```{r}
#| label: count-check
sum(philly_tracts$n_stations)
nrow(stations)
```

**These probably don't match. Discuss as a team:** Why not? Find at least one station that explains the gap.

*Hints:* Look for stations where `GEOID` is `NA` in `stations_with_tract`. For double counts, look for stations that intersect more than one tract: `which(lengths(st_intersects(stations, philly_tracts)) > 1)`.

---

## Part 4: How far is the nearest station?

`st_distance()` returns a **matrix**, not a column. Look at its shape first:

```{r}
#| label: distance-shape
centroids <- st_centroid(st_geometry(philly_tracts))

dim(st_distance(centroids, stations))
```

Instead, find the nearest station for each tract, then measure only that one distance:

```{r}
#| label: nearest
nearest_idx <- st_nearest_feature(centroids, ___)

philly_tracts <- philly_tracts %>%
  mutate(
    dist_ft = as.numeric(
      st_distance(centroids, stations[nearest_idx, ], by_element = ___)
    ),
    dist_mi = dist_ft / ___
  )

summary(philly_tracts$dist_mi)
```

---

## Part 5: Buffers, filters, and cutting the dough

Build a **half-mile** zone around every station and merge the circles into one shape. Remember your units.

```{r}
#| label: buffer
station_zone <- stations %>%
  st_buffer(dist = ___) %>%
  st_union()
```

**Pick whole cookies:** which tracts don't touch the zone at all?

```{r}
#| label: filter-disjoint
outside_zone <- philly_tracts %>%
  st_filter(station_zone, .predicate = ___)

nrow(outside_zone)
```

**Cut the dough:** what share of each tract's area falls inside the zone?

```{r}
#| label: intersection
philly_tracts <- philly_tracts %>%
  mutate(tract_area = as.numeric(st_area(.)))

covered <- philly_tracts %>%
  st_intersection(station_zone) %>%
  mutate(covered_area = as.numeric(st_area(.))) %>%
  st_drop_geometry() %>%
  select(GEOID, covered_area)

philly_tracts <- philly_tracts %>%
  left_join(covered, by = "GEOID") %>%
  mutate(pct_covered = replace_na(covered_area / tract_area, 0) * 100)
```

**Notice:** that last step is the `left_join()` from Part 1b. It's a join by *what* (GEOID), used to bring a join-by-*where* result back into the table. The two kinds of join work together.

---

## Part 6: The Challenge

You now have three different definitions of "has access":

1. **Distance:** the tract centroid is within X miles of a station (`dist_mi`)
2. **Touch:** the tract touches the half-mile zone at all (`st_filter`)
3. **Coverage:** at least Y% of the tract's area is inside the zone (`pct_covered`)

### Your team's task

1. **Define "low-income."** For example, you could use the bottom quartile of tract median income, or a dollar threshold. Pick one and write it down.
2. **Define "access."** Choose one of the three definitions above and set its threshold.
3. **Answer the question.** Which low-income tracts lack access? Produce:
   - **The count** of low-income tracts without access
   - **A map** showing them (a `geom_sf()` layer plus the stations)
   - **The reliability breakdown:** of those tracts, how many have High / Moderate / Low reliability income estimates?
4. **Defend one decision** in 2–3 sentences. Why this access definition and not another? How would your count change if you used a different one?

```{r}
#| label: challenge
# Your team's code here



```

### Scoring (for bragging rights, not grades)

| | Points |
|---|---|
| Count is correct *for your stated definitions* | 2 |
| Map shows the right tracts and the stations, with a title that states your definitions | 2 |
| Reliability breakdown included | 1 |
| Randomly chosen presenter explains the decision clearly | 3 |
| Team shows how the count changes under a second definition | +1 |

### Before you present, discuss

- If a tract is "low-income" with a CV of 45%, does it belong on your list? What would you tell Indego about it?
- Your three access definitions give different counts. Which one would an Indego planner prefer? Which one would a neighborhood advocate prefer?

---

## If you're curious 

- Indego stations have a `totalDocks` property. Does "access" mean the same thing next to a 10-dock station and a 35-dock station?
- Replace the centroid with `st_point_on_surface()`. Does any tract's distance change a lot? Why might a centroid fall in an odd place for some tract shapes?
- Straight-line distance ignores the Schuylkill, I-95, and rail yards. Which tracts do you suspect are *farther* in practice than your map shows?
