---
title: "Assignment3_Verify the Claim: Philadelphia Poverty Change"
subtitle: "Testing if ACS data supports the 2018 news story"
author: Gautami Kasat
date: today
format:
  html:
    toc: true
    code-fold: true
    code-summary: "View Code"
    embed-resources: true
execute:
  warning: false
  message: false
---

## The Enquiry

This report tests whether the ACS data supports the claims in Alfred Lubrano's 2018 article, ["New Census figures on Philly neighborhoods show inequality, high numbers of whites living in poverty"](https://whyy.org/articles/new-census-figures-on-philly-neighborhoods-show-inequality-high-numbers-of-whites-living-in-poverty/) (WHYY / Philadelphia Inquirer, December 6, 2018). 

## Setup

```{r}
#| include: false
library(tidyverse)
library(tidycensus)
library(sf)
```

## Part 1: Reconstruct

### Variables 
I pulled four ACS variables. For overall poverty, B17001_002 counts people below the poverty level in the past twelve months, and B17001_001 gives the population for whom poverty status is determined. Because several claims are specifically about white poverty, I also pulled B17001A_001 and B17001A_002, the same table restricted to people identifying as white alone. ACS publishes counts, not rates, so both the rate and its margin of error have to be calculated. I used the 2006-2010 and 2013-2017 five-year estimates at the census tract level. Five-year estimates are the only ones that reach tracts; one-year estimates cover only areas of 65,000 people or more.

### Assumptions 
The article does not document its method, so several choices are mine. It does not name the ACS table it used; I assumed B17001. It says "white" without defining it, so I used B17001A, white alone, which includes people who also identify as Hispanic. And it reports neighborhood figures, which are not Census geographies: its reporters aggregated tracts into boundaries of their own choosing and never published which tracts went where, so my numbers will not match theirs. Every figure in the article is a point estimate, with no margins of error given. 

### Overlap
The two periods do not overlap. Census Bureau guidance is explicit that only non-overlapping five-year estimates should be compared over time, so on this point the article is on solid ground.

### Pulling the ACS data: 
```{r}
pov_2010 <- get_acs(
  geography = "tract",
  variables = c( pov_total = "B17001_001",pov_poor = "B17001_002", wht_total = "B17001A_001",wht_poor = "B17001A_002"),
  state = "PA",
  county = "Philadelphia",
  year = 2010,
  output = "wide",
  survey = "acs5"
)

pov_2017 <- get_acs(
  geography = "tract",
  variables = c( pov_total = "B17001_001",pov_poor = "B17001_002", wht_total = "B17001A_001",wht_poor = "B17001A_002"),
  state = "PA",
  county = "Philadelphia",
  year = 2017,
  output = "wide",
  geometry = TRUE,
  survey = "acs5"
)

setdiff(pov_2010$GEOID, pov_2017$GEOID)
```
